---
title: "Winter_project"
author: "Oxana Kolpakova"
date: "06 04 2022"
output:
  pdf_document: default
  html_document: default
---
Load the necessary libraries. 
```{r message=FALSE, include=FALSE}
library(tidyverse)
library(readxl)
library(caret)
library(factoextra)
library(psych)
library(vegan)
library(caTools)
library(skimr)
library(zoo)
library(lsr)
```
Import data file
```{r}
micedata <- read.csv ("C:/Users/1/Desktop/Data_Cortex_Nuclear.csv", sep =";", header = T, na.strings = c(" ", "NA"))
```
Data correction.
```{r message=FALSE, include=FALSE}
head(micedata)
str(micedata)

#Change protein level colunns  to numeric and 
#"Genotype", "Treatment", "Behavior", "class" column to factor 

micedata[, c(5, 6,9,11,19,25,26,33,35,37,47,64,65,67,68,78)] <- sapply(micedata[, c(5, 6,9,11,19,25,26,33,35, 37, 47,64,65,67,68,78)], as.numeric)
micedata[, c(79:82)] <- lapply(micedata[, c(79:82)], factor)

#change strange name to normal
names(micedata)[names(micedata) == 'п.їMouseID'] <- 'MouseID'
str(micedata)  
```
1. Data description.

-72 mice were used in the experiment.
-The data contains: expression level of 77 proteins, 2 Genotypes, 2 Treatments, 2 Behavior, 8 class.
-The class data is balanced.
-Data for 23 proteins are complete without NA, for 5 proteins about 15% is lost, for the rest about 5% is lost 
```{r}
str(micedata)
skim(micedata)  
#  balance  of class
prop.table(table(micedata$class))
```
2. Are there differences in the level of BDNF_N production depending on the class in the experiment?
ANOVA analysis of variance showed that the expression level of BDNF_N was class dependent, p-value < 0.001. With the help of the TukeyHSD test, we found out which classes there are differences between. Effect Size eta-Squared = 0.110
The expression level of BDNF_N differs significantly for the classes:
c-SC-m vs. c-CS-m
c-SC-m vs. c-CS-s
t-SC-s vs. -c-SC-m
t-CS-s vs. c-CS-s
t-CS-s vs. c-CS-m
t-SC-m vs. c-SC-m
t-CS-m vs. c-CS-s
c-SC-s vs. c-CS-s
t-CS-m vs. c-CS-m
c-SC-s vs. c-CS-m
c-SC-s vs. c-SC-m
t-CS-m vs. c-SC-m
t-SC-m vs. c-CS-s
t-SC-s vs. t-CS-s
t-SC-m vs.c-CS-m
```{r message=TRUE}
#replece NA to column mean
micedata_NA <- replace(micedata, TRUE, lapply(micedata, na.aggregate))

fit <- aov(BDNF_N ~ class, data=micedata_NA) 
summary(fit)

#differences between classes
TukeyHSD(fit)

#Effect Size eta-Squared = 0.110
etaSquared(fit, anova = TRUE)

means <- tapply(micedata_NA$BDNF_N, micedata_NA$class, mean)
sd <- tapply(micedata_NA$BDNF_N, micedata_NA$class, sd)
class <- c("c-CS-m", "c-CS-s", "c-SC-m", "c-SC-s", "t-CS-m", "t-CS-s", "t-SC-m", "t-SC-s")
plot_data <- data.frame(means, sd, class) 

plot_data

ggplot(plot_data, aes(x=class, y=means)) + 
    geom_bar(stat="identity", width=.4, fill="gold") +
    geom_errorbar(aes(ymin=means-sd, ymax=means+sd), width=.2) +
    theme_bw()
```

3. Build a linear model for predicting  of the ERBB4_N  expression level based on data on other proteins in the experiment. 

Prepare data for lm analysis
```{r message=FALSE}
#Prepare data for analysis

#Remove columns non containing protein levels and convert to matrix.
proteins <- as.matrix(micedata[,c(-1, -79:-82)])

# z-score data
proteins.norm <- apply(proteins, 2, function(x){
  (x-mean(x, na.rm = T))/diff(range(x, na.rm = T))})

# Substitute NAs with mean values
proteins.norm <- apply(proteins.norm, 2, function(x){
  x[is.na(x)] <- mean(x, na.rm = TRUE)
  x
  })

#Divide data into training and test samples.

set.seed(101) 
# create a vector with randomly selected observation numbers 
tr.index = sample(1:nrow(proteins.norm), nrow(proteins.norm)*0.8) 
 
#form a training sample,80%
trSet = proteins.norm[tr.index, ]
#form a control sample, 20%
testSet = proteins.norm[-tr.index, ]

trSet1 <- as.data.frame(trSet, row.names = NULL, optional = FALSE)
testSet1 <- as.data.frame(testSet, row.names = NULL, optional = FALSE)
```
Make a linear model that predicts the ERBB4_N expression level by  all available predictors, Adjusted R-squared=0.8028873, the model is not bad. This is also visible on the plots.

```{r message=TRUE}

model <- lm(ERBB4_N ~ ., data=trSet1)
#summmary(model)
summary(model)
summary(model)$adj.r.squared

result <- predict(model, newdata = testSet1)
result

par(mfrow = c(2, 2))
plot(model)

```
4. Run and visualise the result of PCA. 8 PCs explain 78% of variance
```{r message=TRUE}
pca <- prcomp(proteins.norm)
summary(pca)#  8 PCs explain 78% of variance

fviz_eig(pca)
 
screeplot(pca, type = "lines", bstick = TRUE) # it is necessary to leave 8PCs 

fviz_pca_ind(pca, col.ind = micedata$Behavior, geom = "point", pch = 20) +
  geom_point(size=2, aes(col = micedata$Behavior)) 

fviz_pca_var(pca, col.var = "contrib")
```

